\section{Expression versus statement}
\label{sec:expression versus statement}
In this book we talk about expressions and statements, but%
\gomarginpar{This section comes from \cite{so_expression_vs_statement}.}
what \emph{is} the difference between the two?
In short:
\begin{description}
\item[Expression] Something which evaluates to a value, like:
\lstinline{1+2/x} ;
\item[Statement] A line of code which does something, like
\lstinline{goto Error} .
\end{description}

The distinction was crystal-clear in the earliest general-purpose
programming languages, like Fortran. In Fortran, a statement was one
unit of execution: "a thing that you did". 
An expression on its own couldn't do anything. You had to assign it to a
variable.
\begin{display}
1 + 2 / X
\end{display}
is an error in Fortran, because it doesn't do anything. You had to do
something with that expression: 
\begin{display}{X = 1 + 2 / X}\end{display}

The earliest popular language to blur the lines was C. The designers of
C realized that no harm was done if you were allowed to evaluate an
expression and throw away the result. In C, every expression could be a
statement: 
\begin{display}1 + 2 / x\end{display}
is a totally legit statement even though absolutely nothing will happen.
Why? Because in C, expressions could have side-effects --- they could
change something: \begin{display}{1 + 2 / callfunc(12)}\end{display}

Because \func{callfunc()} might just do something useful.
Once you allow any expression to be a statement, you might as well allow
the assignment operator (=) inside expressions. That's why C lets you do
things like: \lstinline{callfunc(x = 2)}.
This evaluates the expression \lstinline{x = 2} (assigning the value of 2 to x) and
then passes that (the 2) to the function \func{callfunc()}.

Luckily this blurring is significantly lessened in Go.
The only Go
expressions that can be statements are function calls, method calls,
and channel operations. It's almost like Fortran. 

Functional languages like Lisp don't have statements.
All they have is expressions. 
